Perturbing SU($n$)-symmetric interactions

### 2020-01-16 collective_sun/sun_perturbation.tex
address the case of a translationally invariant lattice with dimension D>1

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{0,1}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-body
operator $O_{pq}$.  If the two-body operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{pk} w_{Okq} + w_{Opk} h_{kq}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ indexed by a scalar $\Delta$ is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
The eigenvalue problem in \eqref{eq:cond_2} can be simplified somewhat
when the interaction Hamiltonian $H_0$ is translationally invariant;
we discuss this simplification in Appendix \ref{sec:trans_inv}.

@@

Here we simplify the eigenvalue problems that appear at second order
in the single- and two-body theory in Sections
\ref{sec:single_body_pert} and \ref{sec:two_body_pert} in the case of
a translationally invariant system with periodic boundary conditions.
We solve the single-body eigenvalue problem analytically, and reduce
the two-body eigenvalue problem to that of diagonalizing an
$\floor{N/2}\times\floor{N/2}$ matrix.

@@

For reference, the single-body eigenvalue problem is
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  \label{eq:cond_1_ref}
\end{align}
where $\m h$ is a matrix of all couplings $h_{pq}$ that appear in the
interaction Hamiltonian $H_0$ in \eqref{eq:ints}; $\v h$ is a vector
of all $h_p\equiv\sum_q h_{pq}$; $\diag\v h$ is a matrix with $\v h$
on the diagonal and zeroes everywhere else; and $\v v$ is a vector of
coefficients used to construct the single-body perturbation
$\V_1\p{\v v}$ defined in \eqref{eq:perturbations}.  When the
interaction Hamiltonian $H_0$ is translationally invariant, the
couplings $h_{pq}$ depend only on the separation $\abs{p-q}$, so the
eigenvectors of $\m h$ are plane waves of the form
\begin{align}
  \v v_k \equiv \sum_p e^{ipk} \ket{p},
\end{align}
where $k\in\ZZ_N\times2\pi/N$.  The corresponding eigenvalues can be
determined by expanding
\begin{align}
  \m h\c\v v_k = \sum_{p,q} h_{pq} e^{iqk} \ket{p}
  = \sum_{p,d} h_{p,d+p} e^{i\p{d+p}k} \ket{p}
  = \sum_d h_{0,d} e^{idk} \sum_p e^{ipk}\ket{p}
  = \sum_d h_{0,d} \cos\p{dk} \v v_k.
\end{align}
The remainder of \eqref{eq:cond_1_ref} that we need to sort out is
$\diag\v h$, where all $h_p=\sum_qh_{pq}=\sum_dh_{0,d}$ are equal,
which implies that $\diag\v h=\sum_dh_{0,d}$ is a scalar.  We thus
find that
\begin{align}
  \p{\m h - \diag\v h}\c\v v_k = \Delta_k \v v_k,
  &&
  \Delta_k \equiv \sum_d h_{0,d} \sp{\cos\p{dk}-1}.
\end{align}
Note that $\Delta_k=\Delta_{-k}$, which implies that there are at most
$\lceil N/2\rceil$ distinct eigenvalues of $\m h - \diag\v h$ (one of
which is $\Delta_0=0$).

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
  \p{2h_{pq}-h_p-h_q} w_{pq}
  + \sum_{k\notin\set{p,q}} \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta w_{pq},
  \label{eq:cond_2_ref_start}
where $h_p\equiv\sum_q h_{pq}$.  If the interaction Hamiltonian $H_0$
and perturbation $\V_2\p{\m w}$ are translationally invariant, then
the couplings $h_{pq}$ and coefficients $w_{pq}$ depend only on the
separation $\abs{p-q}$.  In this case, all $h_p=h_0$ are equal, and
\eqref{eq:cond_2_ref_start} can be written in the form
  2 \m h\circ \m w + \m h\c\m w + \m w\c\m h
  = \sp{\Delta + 2h_0} \m w,
  \label{eq:cond_2_ref}
where $\m h\circ\m w$ denotes an element-wise (Kronecker) product of
$\m h$ and $\m w$, i.e.~with matrix elements $h_{pq}w_{pq}$.  In order
to find solutions to \eqref{eq:cond_2_ref}, we we expand the matrices
$\m h,\m w$ in a basis that respects translational symmetry.  To this
end, for all $d\in\ZZ_N$ we define the {\it shift matrices}
  \m s_d \equiv \sum_p \op{p+d}{p},
where the sum $p+d$ is implicitly taken modulo $N$.  The shift
matrices allow us to expand
  \m h = \sum_{p,q} h_{pq} \op{p}{q}
  = \sum_{d,q} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m s_d,
  \m w = \sum_d w_{d,0} \m s_d,
and write the condition in \eqref{eq:cond_2_ref} as
  \sum_{d\ne 0} h_{d,0} w_{d,0} \m s_d
  + \sum_{\substack{c,d\ne0\\c+d\ne0}} h_{c,0} w_{d,0} \m s_{c+d}
  = \p{\f12\Delta + h_0} \sum_{d\ne 0} w_{d,0} \m s_d,
where we explicitly exclude terms that always vanish due to the fact
that $\m h,\m w$ are zero on the diagonal.  With some re-indexing, we
can write
  \sum_{d\ne 0} h_{d,0} w_{d,0} \m s_d
  + \sum_{\substack{c,d\ne0}} h_{d-c,0} w_{c,0} \m s_d
  = \p{\f12\Delta + h_0} \sum_{d\ne 0} w_{d,0} \m s_d,
  \label{eq:cond_2_shift}
and define
  \ut{\check{\m h}}
  \equiv \sum_{d\ne0} h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{d-c,0} \op{d}{c},
  \ut{\v{\m w}} \equiv \sum_{d\ne0} w_{d,0} \ket{d},
in order to express \eqref{eq:cond_2_shift} as the eigenvalue
equation
  \ut{\check{\m h}} \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.
  \label{eq:cond_2_square}
We can reduce this eigenvalue problem even further by making use of a
reflection symmetry that comes from the fact that $h_{pq}=h_{qp}$ and
$w_{pq}=w_{qp}$, which implies $h_{d,0}=h_{-d,0}$ and
$w_{d,0}=w_{-d,0}$ (where all indices are taken modulo $N$, taking
$-d\to N-d$).  To this end, we define a complete basis of orthogonal
vectors that are symmetric under reflection, namely for all
$d\in\VV_N\equiv\set{1,2,\cdots,\floor{N/2}}$,
  \ket{d_+} \equiv \f{\eta_d}{\sqrt{2}}\p{\ket{d} + \ket{-d}},
  \eta_d \equiv
  \begin{cases}
    1 & d \ne N/2, \\
    1/\sqrt{2} & d = N/2
  \end{cases},
where the special case of $d=N/2$ affects only even $N$, for which
$\ket{N/2}=\ket{-N/2}$.  We can use these vectors to define a
projector onto the reflection-symmetric subspace and expand
$\ut{\v{\m w}}$ as
  \P_+ \equiv \sum_{d\in\VV_N} \op{d_+},
  \ut{\v{\m w}} = \sum_{d\in\VV_N} w_{d,0} \,
  \eta_d \sqrt{2} \ket{d_+}.
Finally, we project $\ut{\check{\m h}}$ onto the reflection-symmetric
subspace,
  \P_+ \ut{\check{\m h}} \P_+
  &= \sum_{a,b\in\VV_N} \sum_{c,d\ne0}
  \p{h_{d,0} \delta_{cd} + h_{c-d,0}} \op{a_+} \op{d} \op{b_+} \\
  &= \sum_{a,b\in\VV_N} \op{a_+}{b_+} \f12 \eta_a\eta_b
  \sum_{c,d\ne0} \p{h_{d,0} \delta_{cd} + h_{c-d,0}}
  \p{\delta_{ac} + \delta_{-a,c}} \p{\delta_{bd} + \delta_{-b,d}} \\
  &= \sum_{a\in\VV_N} h_{a,0} \op{a_+}
  + \sum_{a,b\in\VV_N} \eta_a\eta_b
  \p{h_{a-b,0} + h_{a+b,0}} \op{a_+}{b_+},
which allows us to write down the reduced eigenvalue problem
  \P_+ \ut{\check{\m h}} \P_+ \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.
Note that the factors of $\eta_a,\eta_b$ in
$\P_+ \ut{\check{\m h}} \P_+$ only matter when $N$ is even, in which
case these factors merely multiply the last row and last column of
$\P_+ \ut{\check{\m h}} \P_+$ by a factor of $1/\sqrt{2}$.

>>>>

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{0,1}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-body
operator $O_{pq}$.  If the two-body operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{pk} w_{Okq} + w_{Opk} h_{kq}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ indexed by a scalar $\Delta$ is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
The eigenvalue problem in \eqref{eq:cond_2} can be simplified further
when the interaction Hamiltonian $H_0$ is translationally invariant;
we discuss this simplification in Appendix \ref{sec:trans_inv}.

@@

Here we simplify the eigenvalue problems that appear at second order
in the single- and two-body theory in Sections
\ref{sec:single_body_pert} and \ref{sec:two_body_pert} in the case of
a translationally invariant system with periodic boundary conditions.
We solve the single-body eigenvalue problem analytically, and reduce
the two-body eigenvalue problem to that of diagonalizing an
$\p{N-1}\times\p{N-1}$ matrix.  Finally, we discuss the reduction of
the two-body problem to that of diagonalizing a
$\sim\p{N/2^D}\times\p{N/2^D}$ matrix, and work out the case of $D=1$
explicitly as an example.

@@

For reference, the single-body eigenvalue problem is
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  \label{eq:cond_1_ref}
\end{align}
where $\m h$ is a matrix of all couplings $h_{pq}$ that appear in the
interaction Hamiltonian $H_0$ in \eqref{eq:ints}; $\v h$ is a vector
of all $h_p\equiv\sum_q h_{pq}$; $\diag\v h$ is a matrix with $\v h$
on the diagonal and zeroes everywhere else; and $\v v$ is a vector of
coefficients used to construct the single-body perturbation
$\V_1\p{\v v}$ defined in \eqref{eq:perturbations}.  When the
interaction Hamiltonian $H_0$ is translationally invariant, the
couplings $h_{pq}$ depend only on the separation $\abs{p-q}$, so the
eigenvectors of $\m h$ are plane waves of the form
\begin{align}
  \v v_k \equiv \sum_p e^{ipk} \ket{p},
\end{align}
where $k\in\ZZ_N\times2\pi/N$ (with
$\ZZ_N\equiv\set{0,1,\cdots,N-1}$).  The corresponding eigenvalues can
be determined by expanding
\begin{align}
  \m h\c\v v_k = \sum_{p,q} h_{pq} e^{iqk} \ket{p}
  = \sum_{p,d} h_{p,d+p} e^{i\p{d+p}k} \ket{p}
  = \sum_d h_{0,d} e^{idk} \v v_k
  = \sum_d h_{0,d} \cos\p{dk} \v v_k,
\end{align}
where the imaginary terms vanish due to the fact that
$h_{0,d}=h_{0,-d}$.  The remainder of \eqref{eq:cond_1_ref} that we
need to sort out is $\diag\v h$, where all
$h_p=\sum_qh_{pq}=\sum_dh_{0,d}$ are equal, which implies that
$\diag\v h=\sum_dh_{0,d}$ is a scalar.  We thus find that
\begin{align}
  \p{\m h - \diag\v h}\c\v v_k = \Delta_k \v v_k,
  &&
  \Delta_k \equiv \sum_d h_{0,d} \sp{\cos\p{dk}-1}.
\end{align}
This solution also holds for a $D$-dimensional periodic lattice of
$N=L^D$ spins, with $d\in\ZZ_L^D$ and $k\in\ZZ_L^D\times 2\pi/L$ (and
$dk\to d\c k\equiv\sum_{j=1}^D d_j k_j$).

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \p{2h_{pq}-h_p-h_q} w_{pq}
  + \sum_{k\notin\set{p,q}} \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta w_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
where $h_p\equiv\sum_q h_{pq}$.  If the interaction Hamiltonian $H_0$
and perturbation $\V_2\p{\m w}$ are translationally invariant, then
the couplings $h_{pq}$ and coefficients $w_{pq}$ depend on $p$ and $q$
only through the separation $\abs{p-q}$.  In this case, all $h_p=h_0$
are equal, and \eqref{eq:cond_2_ref_start} can be written in the form
\begin{align}
  2 \m h\circ \m w + \m h\c\m w + \m w\c\m h
  = \sp{\Delta + 2h_0} \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h\circ\m w$ denotes an element-wise (Kronecker) product of
$\m h$ and $\m w$, i.e.~with matrix elements $h_{pq}w_{pq}$.  In order
to find solutions to \eqref{eq:cond_2_ref}, we we expand the matrices
$\m h,\m w$ in a basis that respects translational symmetry.  To this
end, for all $d\in\ZZ_N$ we define the {\it shift matrices}
\begin{align}
  \m s_d \equiv \sum_p \op{p+d}{p},
\end{align}
where the sum $p+d$ is implicitly taken modulo $N$.  The shift
matrices allow us to expand
\begin{align}
  \m h = \sum_{p,q} h_{pq} \op{p}{q}
  = \sum_{d,q} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m s_d,
  &&
  \m w = \sum_d w_{d,0} \m s_d,
\end{align}
and write the condition in \eqref{eq:cond_2_ref} as
\begin{align}
  \sum_{d\ne 0} h_{d,0} w_{d,0} \m s_d
  + \sum_{\substack{c,d\ne0\\c+d\ne0}} h_{c,0} w_{d,0} \m s_{c+d}
  = \p{\f12\Delta + h_0} \sum_{d\ne 0} w_{d,0} \m s_d,
\end{align}
where we explicitly exclude terms that always vanish due to the fact
that $\m h,\m w$ are zero on the diagonal.  With some re-indexing, we
can write
\begin{align}
  \sum_{d\ne 0} h_{d,0} w_{d,0} \m s_d
  + \sum_{\substack{c,d\ne0}} h_{d-c,0} w_{c,0} \m s_d
  = \p{\f12\Delta + h_0} \sum_{d\ne 0} w_{d,0} \m s_d,
  \label{eq:cond_2_shift}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}
  \equiv \sum_{d\ne0} h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{d-c,0} \op{d}{c},
  &&
  \ut{\v{\m w}} \equiv \sum_{d\ne0} w_{d,0} \ket{d},
\end{align}
in order to express \eqref{eq:cond_2_shift} as the eigenvalue
equation
\begin{align}
  \ut{\check{\m h}} \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.
  \label{eq:cond_2_square}
\end{align}
This eigenvalue problem takes an identical form a $D$ dimensional
periodic lattice of $N=L^D$ spins, in which case lattice sites and
shift matrices are indexed by vectors $d\in\ZZ_L^D$.

We can reduce the eigenvalue problem in \eqref{eq:cond_2_square} even
further by making use of reflection symmetries such as
$h_{d,0}=h_{-d,0}$.  In $D$ dimensions, there is one such symmetry for
each coordinate that indexes lattice sites, which implies that the
degrees of freedom in $\ut{\v{\m w}}$ gets reduced by a factor of
$\sim 2^D$.  Reflection symmetries similarly reduce the degrees of
freedom in $\ut{\check{\m h}}$ from $\sim N\times N$ to
$\sim\p{N/2^D}\times\p{N/2^D}$.  We work out the case of $D=1$ below
as an example, and leave the general case of arbitrary dimension
$D\ge1$ to the future, if and when it becomes necessary.

In order to exploit reflection symmetry in one dimension ($D=1$), we
define a complete basis of orthogonal vectors that are symmetric under
reflection: for all $d\in\VV_N\equiv\set{1,2,\cdots,\floor{N/2}}$,
\begin{align}
  \ket{d_+} \equiv \f{\eta_d}{\sqrt{2}}\p{\ket{d} + \ket{-d}},
  &&
  \eta_d \equiv
  \begin{cases}
    1 & d \ne N/2, \\
    1/\sqrt{2} & d = N/2
  \end{cases},
\end{align}
where the special case of $d=N/2$ affects only even $N$, for which
$\ket{N/2}=\ket{-N/2}$.  We can use these vectors to define a
projector onto the reflection-symmetric subspace and expand
$\ut{\v{\m w}}$ as
\begin{align}
  \P_+ \equiv \sum_{d\in\VV_N} \op{d_+},
  &&
  \ut{\v{\m w}} = \sum_{d\in\VV_N} w_{d,0} \,
  \eta_d \sqrt{2} \ket{d_+}.
\end{align}
Finally, we project $\ut{\check{\m h}}$ onto the reflection-symmetric
subspace,
\begin{align}
  \P_+ \ut{\check{\m h}} \P_+
  &= \sum_{a,b\in\VV_N} \sum_{c,d\ne0}
  \p{h_{d,0} \delta_{cd} + h_{c-d,0}} \op{a_+} \op{d} \op{b_+} \\
  &= \sum_{a,b\in\VV_N} \op{a_+}{b_+} \f12 \eta_a\eta_b
  \sum_{c,d\ne0} \p{h_{d,0} \delta_{cd} + h_{c-d,0}}
  \p{\delta_{ac} + \delta_{-a,c}} \p{\delta_{bd} + \delta_{-b,d}} \\
  &= \sum_{a\in\VV_N} h_{a,0} \op{a_+}
  + \sum_{a,b\in\VV_N} \eta_a\eta_b
  \p{h_{a-b,0} + h_{a+b,0}} \op{a_+}{b_+},
\end{align}
which allows us to write down the reduced eigenvalue problem
\begin{align}
  \P_+ \ut{\check{\m h}} \P_+ \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.
\end{align}
Note that the factors of $\eta_a,\eta_b$ in
$\P_+ \ut{\check{\m h}} \P_+$ only matter when $N$ is even, in which
case these factors merely multiply the last row and last column of
$\P_+ \ut{\check{\m h}} \P_+$ by a factor of $1/\sqrt{2}$.
